\documentclass[11pt,letterpaper]{article}

% ── Fonts ─────────────────────────────────────────────────────────
\usepackage[T1]{fontenc}
\usepackage{newtxtext}
\usepackage{amssymb}
\usepackage{newtxmath}
\usepackage[scaled=0.85]{inconsolata}

% ── Layout ────────────────────────────────────────────────────────
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\setstretch{1.08}
\usepackage{parskip}
\setlength{\parskip}{0.5\baselineskip plus 2pt minus 1pt}

% ── Section headings ──────────────────────────────────────────────
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection}{1em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.8em}{}
\titleformat{\subsubsection}{\normalsize\itshape}{\thesubsubsection}{0.8em}{}
\titlespacing*{\section}{0pt}{1.8ex plus 0.5ex}{0.8ex plus 0.2ex}
\titlespacing*{\subsection}{0pt}{1.4ex plus 0.3ex}{0.4ex plus 0.1ex}

% ── Math ──────────────────────────────────────────────────────────
\usepackage{amsmath}
\usepackage{mathtools}
\let\openbox\relax  % newtxmath already defines \openbox; let amsthm redefine it
\usepackage{amsthm}

% ── Tables ────────────────────────────────────────────────────────
\usepackage{booktabs}
\usepackage{array}
\usepackage{tabularx}
\renewcommand{\arraystretch}{1.15}
\usepackage{caption}
\captionsetup{font=small, labelfont=bf, labelsep=period, skip=6pt}

% ── Floats & graphics ─────────────────────────────────────────────
\usepackage{float}
\usepackage{graphicx}
\usepackage{placeins}

% ── Lists ─────────────────────────────────────────────────────────
\usepackage{enumitem}
\setlist{nosep, leftmargin=1.5em}

% ── Colors & links ────────────────────────────────────────────────
\usepackage{xcolor}
\definecolor{linkblue}{HTML}{0B5394}
\definecolor{citegreen}{HTML}{1B7340}
\definecolor{urlgray}{HTML}{444444}
\usepackage[
  colorlinks=true, linkcolor=linkblue, citecolor=citegreen, urlcolor=urlgray,
  bookmarksnumbered=true, pdfstartview=FitH,
]{hyperref}
\usepackage[noabbrev,capitalise]{cleveref}
\usepackage{microtype}

% ── Theorem-like environments ─────────────────────────────────────
\theoremstyle{definition}
\newtheorem{estimand}{Estimand}
\newtheorem{measure}{Measure}
\newtheorem{definition}{Definition}[section]
\newtheorem{remark}{Remark}[section]
\theoremstyle{plain}
\newtheorem{hypothesis}{Hypothesis}
\newtheorem{gate}{Gate}

% explicit cref names for the custom theorem types (singular / plural / range)
\crefname{estimand}{Estimand}{Estimands}
\crefname{measure}{Measure}{Measures}
\crefname{hypothesis}{Hypothesis}{Hypotheses}
\crefname{gate}{Gate}{Gates}
\crefname{definition}{Definition}{Definitions}
\crefname{remark}{Remark}{Remarks}

\newcommand{\Vstruct}{V^{\mathrm{struct}}}
\newcommand{\Vlat}{V^{\mathrm{lat}}}
\newcommand{\Real}{\mathbb{R}}
\newcommand{\card}[1]{\lvert #1 \rvert}

\title{\textbf{An Empirical Specification for the $V$-Extension}\\[2pt]
\large Measuring Per-Capita Structural Novelty to Test Core Claim~6\\
\normalsize (the companion instrument to the WS3 $C$--$V$ primer)}
\author{Originality Program --- Whitespace 2 $\to$ Whitespace 3 bridge}
\date{July 2026}

\begin{document}
\maketitle
\tableofcontents
\clearpage

%==============================================================
\section{Purpose, placement, and the one-sentence claim}
%==============================================================

The WS3 primer (\texttt{cv-reconciliation-primer.tex}) defines two output gauges,
cumulative complexity $C(t)$ and per-capita variance generation $V(t)$, and shows
that Whitespace~2 (WS2) pinned the collective side ($C$-family breadth $W$,
canonical concentration $H$, demographic diversity $D$; all rising) but left the
per-capita gauge $V$ \emph{unmeasured}. \textbf{Core Claim~6} of that primer is
therefore a prediction rather than a result:
\begin{quote}
per-capita \emph{structural} novelty $\Vstruct(t)$ is declining---strongest where
the canon is most concentrated (Physics)---while per-capita \emph{topical}
novelty $\Vlat(t)$ need not decline.
\end{quote}
This document specifies the empirical instrument that converts Core Claim~6 into
a falsifiable test, reusing data already in hand.

\paragraph{Placement.} Operationally this is a \textbf{WS2 Phase~2.4} (a
per-capita extension of the Phase~2.2/2.3 machinery); scientifically it is the
\textbf{empirical anchor for WS3}. It requires \emph{no new embedding}: the base-1M
SciNCL, Qwen3, and SPECTER2 vectors, the 1M source corpus (references +
concepts + authorships), the Phase-1.3 demographic substrate, and the
reference-canonicity machinery all already exist. Marginal compute is a citation-graph
inversion and panel regressions---\$0 (one optional robustness pull excepted,
\S\ref{sec:feasibility}).

\paragraph{The one design principle.} Everything below is built to respect the
same distinction the theory turns on (primer Remark~2.3): an author's output has
a \emph{content} channel (what they work on $\to \Vlat$) and an \emph{attachment}
channel (what they build on / cite $\to \Vstruct$). The instrument must measure
the two \emph{separately} and per capita, and---per the hard-won WS2 cautions---must
measure the attachment channel in at least one \emph{embedding-free} way, because
WS2 showed embedding-based semantic trends are model-dependent.

%==============================================================
\section{Estimands: from model quantities to observables}\label{sec:estimands}
%==============================================================

Fix a study window (1970--2023, trimming the incomplete final year). Index papers
$p$, authors $a$, years $t$; let $T_p$ be paper $p$'s team size (author count) and
$F(p)$ its field/subfield.

\begin{estimand}[Per-capita structural novelty]\label{est:struct}
For author $a$ in year-window $t$, $\Vstruct_a(t)$ is the mean over $a$'s papers
in $t$ of \emph{persisting attachment-deviant novelty}: a paper contributes when
it (i) builds on non-canonical antecedents (low empirical canon-alignment
$\hat\gamma$, \S\ref{sec:struct}) and (ii) that move \emph{persists}
(\S\ref{sec:persist}). Its field-year aggregate $\hat\Vstruct(t)$ is the
per-capita mean over authors active in $(F,t)$. The estimand of interest is the
\textbf{trend} $\mathrm{d}\hat\Vstruct/\mathrm{d}t$ and its \textbf{gradient in
canon concentration} $H$.
\end{estimand}

\begin{estimand}[Per-capita topical novelty --- the contrast]\label{est:lat}
$\Vlat_a(t)$ is the analogous quantity built from \emph{content} novelty
(distance from topical precedent, \S\ref{sec:lat}) rather than attachment
deviance. Its trend is the \textbf{control contrast}: Core Claim~6 predicts
$\Vstruct$ falls while $\Vlat$ does not.
\end{estimand}

\begin{remark}[Why the two must be measured with the right lens]
WS2 established that \emph{topical} embeddings (SciNCL, Qwen3) track the content
channel and a \emph{citation-tuned} embedding (SPECTER2) tracks the attachment
channel. Accordingly $\Vlat$ is measured on SciNCL/Qwen3 and $\Vstruct$ on
SPECTER2 \emph{and} on embedding-free reference statistics. Using the wrong lens
would re-run the exact confusion WS2 diagnosed.
\end{remark}

%==============================================================
\section{Data inventory (all existing)}
%==============================================================

\begin{table}[h]
\centering
\small
\begin{tabularx}{\textwidth}{@{}lX@{}}
\toprule
Artifact & Contents / role \\
\midrule
base-1M vectors (SciNCL, Qwen3, SPECTER2) & $902{,}531\times768$, row-aligned to
\texttt{metadata.parquet}; SciNCL/Qwen3 $\to\Vlat$, SPECTER2 $\to\Vstruct$
(embedding arm) \\
1M source corpus & \texttt{referenced\_works\_json} (attachment / citations),
\texttt{concepts\_json} (topics, subfields), \texttt{authorships} \\
demographic substrate (Phase 1.3) & per-author \texttt{author\_id},
\texttt{min\_year} (career stage), \texttt{primary\_country}, gender; ORCID flags
for a well-disambiguated subset \\
canon machinery & \texttt{reference\_canonicity}: per-year reference-target
frequencies, Gini $H$, top-$N$ set $\to$ the canon $K(t)$ for $\hat\gamma$ \\
K=50 clusters + subfield maps (Phase 2.3) & data-driven and concept-level
subfields for the $H$-gradient test \\
in-sample citation graph & invert \texttt{referenced\_works\_json} across the
corpus $\to$ time-resolved forward uptake (the persistence filter) \\
\bottomrule
\end{tabularx}
\caption{Inputs. Nothing here requires new embedding or new pulls; the only
derived object is the in-sample forward-citation graph (an edge inversion).}
\label{tab:inputs}
\end{table}

%==============================================================
\section{Measurement}
%==============================================================

\subsection{Topical novelty $\to \Vlat$}\label{sec:lat}

\begin{measure}[Distance-to-precedent (content)]\label{meas:lat}
For paper $p$ published in year $t_p$, field $F$, with topical embedding
$x_p$ (SciNCL; Qwen3 as robustness), define
\begin{equation}
  n^{\mathrm{lat}}(p) \;=\; 1 - \max_{q\in\mathcal{P}(p)} \cos\!\big(x_p, x_q\big),
  \qquad
  \mathcal{P}(p) = \{q : t_q < t_p,\ F(q)=F\}\ \text{(precedent set)},
\end{equation}
optionally averaged over the $\kappa$-nearest precedents rather than the single
max (robustness). Higher $=$ more topically novel relative to prior work. For
tractability $\mathcal{P}(p)$ is a seeded random sample (fixed size) of prior
in-field papers; report the sample size.
\end{measure}

\subsection{Structural novelty $\to \Vstruct$ (three operationalizations)}\label{sec:struct}

The attachment channel is measured \emph{three} ways---two embedding-free, one
embedding---so the headline cannot hinge on any single lens (the WS2 caution).

\begin{measure}[Off-canon reference share --- the empirical $1-\gamma$]\label{meas:offcanon}
Let $R(p)$ be $p$'s reference set and $K(t_p)$ the canon as of $t_p$ (top-weighted
works using \emph{only} data up to $t_p$; \S\ref{sec:controls} on leakage). The
empirical canon-alignment is $\hat\gamma(p) = \card{R(p)\cap K(t_p)}/\card{R(p)}$
(the estimand of primer Eq.~6), and
\begin{equation}
  n^{\mathrm{struct}}_{\mathrm{oc}}(p) \;=\; 1 - \hat\gamma(p)
  \quad\in[0,1].
\end{equation}
This is the most direct image of the model's structural-deviance quantity: how
much of $p$'s foundation lies off the shared canon.
\end{measure}

\begin{measure}[Reference atypicality --- novel recombination (Uzzi-style)]\label{meas:atyp}
For each unordered pair of references $\{u,v\}\subseteq R(p)$, compute the
$z$-score of their observed co-reference frequency against a degree-preserving
null (co-reference expected from each work's citation frequency). Summaries per
paper: the \emph{median} $z$ (conventionality) and the \emph{10th-percentile}
$z$ (the atypical tail). Structural novelty
$n^{\mathrm{struct}}_{\mathrm{atyp}}(p)$ is decreasing in median $z$ / increasing
in a left (atypical) tail. Embedding-free; captures \emph{recombination} novelty
the off-canon share misses.
\end{measure}

\begin{measure}[Citation-geometry distance (SPECTER2)]\label{meas:s2}
With SPECTER2 (citation-tuned) embedding $y_p$, define
$n^{\mathrm{struct}}_{\mathrm{emb}}(p) = 1 - \max_{q\in\mathcal{P}(p)} \cos(y_p,y_q)$
exactly as in \cref{meas:lat} but on the attachment-tuned family. This is the
embedding arm; its agreement or disagreement with the two reference-based
measures is itself a headline diagnostic.
\end{measure}

\subsection{The persistence filter --- the ``$\ge k$ generations'' gate}\label{sec:persist}

A novelty counts only if it \emph{sticks} (primer \cref{est:struct}); a one-off
attachment nobody adopts is churn, not $V$.

\begin{measure}[Uptake-within-window persistence]\label{meas:persist}
Using the in-sample forward-citation graph, let $u_w(p)$ be $p$'s forward
citations accrued within a fixed $w$-year post-publication window (e.g.\ $w=5$),
\emph{age-restricted} so every paper has a full window (drop papers whose window
extends past the snapshot). Define the persistence weight
$s(p) = \mathbf{1}[u_w(p) \ge \theta_w]$ or a smooth $s(p) =
u_w(p)/(u_w(p)+\bar u_w)$, where thresholds/scales are computed \emph{within}
(field, publication-year) cells to remove field/era base-rate differences.
\end{measure}

\begin{remark}[Persistence, not the CD-index]
Per WS3's own pre-commitment (``frame $V$ around any novelty metric, not
specifically CD-index''), persistence is a robust forward-uptake filter, not a
disruption index. A CD/disruption variant may be reported as a \emph{secondary}
column with the Petersen--Holst caveats attached, never as the headline.
\end{remark}

\subsection{Per-capita aggregation and team size}\label{sec:percap}

\begin{measure}[Per-paper persisting novelty, and per-capita roll-up]\label{meas:rollup}
Per-paper persisting structural novelty is
$\nu^{\mathrm{struct}}(p) = n^{\mathrm{struct}}(p)\cdot s(p)$ (analogously for
lateral). Two per-capita estimands:
\begin{enumerate}[label=(\roman*)]
  \item \textbf{Paper-level with team-size control} (primary): analyze
  $\nu^{\mathrm{struct}}(p)$ with $\log T_p$ as a covariate.
  \item \textbf{Fractional per-author} (robustness): attribute
  $\nu^{\mathrm{struct}}(p)/T_p$ to each author, then average within
  $(a,t)$ to form the author panel $\Vstruct_a(t)$.
\end{enumerate}
Because mean team size rose sharply over the window---and larger teams are less
disruptive per capita (WWE)---team size is both a \emph{confound to control} and
\emph{part of the mechanism}; \cref{hyp:withinteam} tests whether the decline
survives \emph{within} team-size strata.
\end{measure}

%==============================================================
\section{Estimator and pre-registered hypotheses}\label{sec:hypotheses}
%==============================================================

The headline estimator is a panel regression of per-paper (or per-author)
persisting novelty on year, with fixed effects and controls:
\begin{equation}\label{eq:panel}
  \nu^{\mathrm{struct}}(p) \;=\; \beta_0 + \beta_1\,t_p
    + \beta_2 \log T_p + \beta_3\,\log(\text{field volume}_{F,t_p})
    + \alpha_{a(p)} + \delta_{F} + \varepsilon_p ,
\end{equation}
with author fixed effects $\alpha_a$ (within-author identification) and field
effects $\delta_F$. Inference by \textbf{permutation} (shuffle $t_p$ within
field) plus the \textbf{absolute} change over the window; \textbf{VIF reported}
for $(t,\log T,\log\text{volume})$; every headline run repeated for
\textbf{$\ge 2$ structural measures} (\cref{meas:offcanon,meas:atyp,meas:s2}).

\begin{hypothesis}[Primary: structural novelty declines]\label{hyp:decline}
$\beta_1 < 0$ for $\nu^{\mathrm{struct}}$, robust across (a) the off-canon share
and the atypicality measure (embedding-free) \emph{and} the SPECTER2 measure, and
(b) permutation significance with a non-negligible absolute effect.
\end{hypothesis}

\begin{hypothesis}[Contrast: topical novelty does not decline]\label{hyp:contrast}
$\beta_1 \ge 0$ or non-significant for $\nu^{\mathrm{lat}}$ (SciNCL and Qwen3).
The dimension-specificity ($\Vstruct\!\downarrow$, $\Vlat$ spared) is the
signature that distinguishes the mechanism from a generic novelty decline.
\end{hypothesis}

\begin{hypothesis}[Differential in canon concentration]\label{hyp:gradient}
Interacting year with subfield canon concentration $H_F$ (from Phase 2.3),
the year$\times H$ coefficient is negative: the structural decline is
\emph{steeper where the canon is more concentrated} (Physics $>$ CS), matching
the WS2 aggregate finding.
\end{hypothesis}

\begin{hypothesis}[Within-team-size robustness]\label{hyp:withinteam}
$\beta_1<0$ survives \emph{within} team-size strata (and with the fractional
per-author estimand), so the decline is not merely mechanical team-size growth.
\end{hypothesis}

\begin{hypothesis}[Mechanism link to $\kappa$]\label{hyp:mech}
Within author (and within team-size), off-canon reference share falls as the
subfield canon concentrates: a direct estimate of the model's conformity acting
on the attachment channel ($\kappa$ suppressing low-$\gamma$ output).
\end{hypothesis}

\paragraph{Negative controls (pre-registered).} (i) Year-shuffle placebo: the
trend must vanish. (ii) Random-canon placebo: recompute $\hat\gamma$ against a
size-matched \emph{random} reference set in place of $K(t)$; the structural trend
must attenuate to null (confirming it is the \emph{canon} that bites, not
reference growth). (iii) A pure-noise novelty series must show no trend.

%==============================================================
\section{Confounds and controls}\label{sec:controls}
%==============================================================

Each control carries a specific WS2/Phase lesson.

\begin{enumerate}[label=\textbf{C\arabic*.}]
  \item \textbf{Citation accrual / age.} Forward-uptake persistence is
  age-confounded (recent papers have less time). Use \emph{fixed $w$-year
  windows}, age-restrict to papers with a complete window, and trim the incomplete
  final years---the Phase-1.4/2.2 age-restriction lesson (raw \texttt{cited\_by\_count}
  is invalid across cohorts).
  \item \textbf{Ratio $\ne$ control.} Report \emph{absolute} per-capita trends for
  $\Vstruct$ and $\Vlat$ \emph{separately}; never divide one novelty channel by
  another (the Phase-2.2 denominator-confound lesson).
  \item \textbf{$\ge 2$ (here 3) structural measures.} Off-canon share +
  atypicality (embedding-free) + SPECTER2 (embedding). A decline seen only in the
  embedding arm is model-specific, not a finding (the SPECTER2/Qwen3 sign-reversal
  lesson).
  \item \textbf{Ill-conditioned $\sigma$.} Judge trends by permutation
  significance \emph{and} absolute magnitude, never a standardized effect on a
  near-flat series (the Phase-2.3 ill-conditioned-$\sigma$ artifact).
  \item \textbf{Collinearity.} Year, $\log$ team size, and $\log$ volume all trend
  together; report VIF and prefer within-stratum comparisons where VIF $\gtrsim 10$
  (the WS-F unidentifiability lesson).
  \item \textbf{Canon leakage / endogeneity.} $K(t_p)$ and the atypicality null use
  \emph{only} data up to $t_p$ (no look-ahead); the canon a paper is scored against
  is the canon it faced.
  \item \textbf{Author disambiguation.} Author FEs and career panels are sensitive
  to OpenAlex split/merge error ($\sim$3\% cross-era mergers, Phase 1.3). Repeat on
  the \textbf{ORCID-backed, career-length-screened} subset as robustness.
  \item \textbf{Survivorship / selection.} Long-observed authors are selected; use
  within-author FEs, entry-cohort controls, and report the estimate on
  \emph{entrants} separately from incumbents.
  \item \textbf{Field composition.} CS/Physics mix and subfield entry shift over
  time; field (and subfield) FEs, plus the per-field decomposition.
\end{enumerate}

%==============================================================
\section{Robustness grid}
%==============================================================

\begin{table}[h]
\centering
\small
\begin{tabularx}{\textwidth}{@{}lX@{}}
\toprule
Axis & Settings \\
\midrule
structural measure & off-canon share / atypicality / SPECTER2 (all three) \\
topical measure & SciNCL / Qwen3 \\
subfield definition & concept level-1 / K=50 clusters (Phase 2.3 pair) \\
persistence window $w$ & \{3, 5, 10\} years (sweep) \\
per-capita form & paper-level + $\log T$ control / fractional per-author \\
disambiguation & full / ORCID-backed subset \\
field & pooled (field FE) / CS-only / Physics-only \\
canon depth $\alpha$ & top-$\{1,5,10\}\%$ for $K(t)$ \\
\bottomrule
\end{tabularx}
\caption{The pre-registered robustness grid. The headline (\cref{hyp:decline})
must hold across the structural-measure and disambiguation axes at minimum; the
gradient (\cref{hyp:gradient}) is read on the field / subfield axes.}
\label{tab:robust}
\end{table}

%==============================================================
\section{Pre-registered evaluation gates}
%==============================================================

\begin{gate}[Instrument validity]
The three structural measures are pairwise positively correlated at the paper
level (they measure a common construct), and the topical measures track the
Phase-2.2 aggregate $W$ trend on a re-aggregation check.
\end{gate}
\begin{gate}[Negative controls pass]
Year-shuffle and random-canon placebos return null; the random-canon placebo
specifically kills the structural trend (isolating the canon as the operative
object).
\end{gate}
\begin{gate}[Headline]
\cref{hyp:decline} holds (structural decline, $\ge 2$ measures, permutation-sig +
material magnitude) \emph{and} \cref{hyp:contrast} holds (topical spared).
Together these confirm Core Claim~6; either failing is reported honestly.
\end{gate}
\begin{gate}[Mechanism + gradient]
\cref{hyp:gradient,hyp:withinteam,hyp:mech} adjudicate whether the decline is
canon-driven, per-capita-real (not team-size-mechanical), and localized in $H$.
\end{gate}

\paragraph{Escape hatches (pre-registered).} If author disambiguation is too
noisy for author FEs even on the ORCID subset, fall back to \emph{team-level}
(paper) panels with cohort controls (diagnose, don't force the author panel---the
Phase-1.3 ``diagnose before escalating'' lesson). If the in-sample forward graph
is too sparse for a $w$-year filter in early decades, widen $w$ or restrict the
persistence-filtered analysis to the post-1985 window and report the unfiltered
novelty trend for the full window alongside.

%==============================================================
\section{Feasibility, cost, and the one data decision}\label{sec:feasibility}
%==============================================================

All measures derive from existing artifacts (\cref{tab:inputs}); the only new
object is the in-sample forward-citation graph (invert
\texttt{referenced\_works\_json} over the corpus---an $O(\text{edges})$ join,
laptop-scale, \$0). Panel regressions are trivial. \textbf{The one decision:} the
persistence filter's forward citations. \emph{Primary (pre-registered):} the
\emph{in-sample} graph---self-contained and \$0, but it undercounts uptake
(only citations from papers inside the 1M sample) and its coverage may drift with
sample density, so persistence thresholds are computed within (field, year) cells
to absorb level shifts. \emph{Optional robustness:} an OpenAlex forward-citation
pull for complete, time-resolved uptake (a bounded API job) to confirm the
in-sample trend is not a coverage artifact. Any bounded coverage caps are
\texttt{log}ged, never silently applied (the no-silent-truncation rule).

%==============================================================
\section{Interpretation matrix: reading outcomes back onto WS3}
%==============================================================

\begin{table}[h]
\centering
\small
\begin{tabularx}{\textwidth}{@{}lX@{}}
\toprule
Outcome & Reading for WS3 (primer Core Claim~6 / \cref{est:struct}) \\
\midrule
$\Vstruct\!\downarrow$, $\Vlat$ spared, gradient in $H$, $\ge 2$ measures agree &
\textbf{Core Claim~6 confirmed on data.} The $C$--$V$ reconciliation's
$\kappa$-on-attachment mechanism is empirically instantiated; WS3 leads with a
measured, not merely modeled, $V$ decline. \\
$\Vstruct$ flat/up despite $H\!\uparrow$ &
\textbf{Falsifies the CD-mechanism reading.} Canon concentration does not suppress
per-capita structural novelty; WS3 must down-weight the canon-deviation $\kappa$
and privilege another mechanism (visible-consensus / attention). \\
decline only in the SPECTER2 arm & model-specific, \emph{not} a finding (repeat of
the WS2 embedding-sensitivity result); report as inconclusive. \\
decline vanishes within team-size strata & the ``small teams disrupt'' pattern is
carried by team-size growth, not a within-stratum conformity effect --- a distinct
(and publishable) refinement of WWE's per-capita story. \\
\bottomrule
\end{tabularx}
\caption{How each outcome feeds back into the WS3 primer. The instrument is
designed so that \emph{every} outcome is informative for the theory---confirming,
falsifying, or relocating the operative $\kappa$ mechanism.}
\label{tab:interp}
\end{table}

\noindent\textbf{One-sentence summary.} Measure, per author/team and per year,
how far scientific output strays from the canonical way of building
(\emph{attachment}, three ways) versus from prior topics (\emph{content}), filter
for what \emph{persists}, and test whether the former---not the latter---declines
and does so most where the canon is strongest: the single measurement that turns
the WS3 primer's Core Claim~6 from a prediction into a result.

\end{document}
